On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-07-23 Code/RankComparison.Rmd
set an example of linewdith 80
updated glmm model and cleaning

@@

clusWilcox.test(Time, group=Year, cluster = Athlete, data = nations, alternative = "greater", method = "ds", exact = TRUE)
```

@@

clusWilcox.test(Time, group=Year, cluster = Athlete, data = comp, alternative = "greater", method = "ds", exact = TRUE)
```

>>>>

Keep the linewidth under 80. Otherwise the code lines are not complete in the
generated pdf files.

@@

clusWilcox.test(Time, group=Year, cluster = Athlete, data = nations,
                alternative = "greater", method = "ds",
                exact = TRUE, B = 10000)
```

@@

clusWilcox.test(Time, group=Year, cluster = Athlete, data = comp,
                alternative = "greater", method = "ds",
                exact = TRUE, B = 10000)
```

### 2023-07-23 Manuscript/manuscript.tex
updated glmm model and cleaning

\usepackage{amsmath, amsfonts}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{cleveref}
\hypersetup{colorlinks = true, linkcolor = blue, citecolor=blue, urlcolor = blue}
\usepackage{natbib}
\usepackage{float}
\usepackage{setspace}
\usepackage{pdfpages}
\usepackage[pagewise]{lineno}
\usepackage{mwe}
\usepackage{comment}
%\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.

\jy{Use double blank lines to separate paragraphs in tex source}

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

\jy{This paragraph reads disconnected. Should it be merged with paragraph~3
  where the study was reviewed?}
Despite their study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is unclear
why IAAF would commission a study only to disregard the results and not
make the recommended changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{Scatterplot2019vs2022}
  \caption{The reaction times from athletes who competed at least once in both 
  the 2019 and 2022 World Track and Field Championships}
  \label{fig:Scatterplot2019vs2022}
\end{figure}
\of{I included this here for now, but feel free to change the location.  I will
add references to it and explanation of it along with some descriptive statistics about
the size and mean/median,std dev, size of the data.}
\jy{Show dot plots which would give the distribution; without uncertainty
  measure, the mean does not tell much.}

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary, 
which makes intuitive sense given the discussion in Section~\ref{sec:dataworld}. 
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
%\eds{we know that many of the athletes compete in multiple world championships,
%  how is this athlete clustering being accounted for here?}
%\jy{We could add an athlete random effect too, if the players are identified in
%  the data.}
%\jy{Owen, do we have athlete id in the data?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time data, the logarithm 
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
 being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

Finally, we remark here that it is also possible to include a heat or race 
\eds{heat? when I first read this, I thought you meant race as in ethnicity}
\of{Yes I think heat or batch effect works better, especially now that I have
removed most talk about heat from the data section} 
\eds{I see, I did not understand this to be Batch effect when I read it the 
first time.  I commented out the Batch text in the Data section - I think it is 
fine to just discuss it here, as it ultimately was not used.  I will defer to 
Dr Yan on this, however.}
effect to account for some heats possibly having faster or slower 
reaction times for everyone involved in the heat.

Although it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other. 
\eds{Did you try it and find that the effect was not needed?  If yes,
I would include that here.  That argument 
would be more compelling for its non-use than its difficulty in interpretation.}
\of{We found that including the race effect resulted in the venue effect not
being significant, and thus Prof Yan suggested that we don't include it.  I am
wondering if this is even worth mentioning? Prof Yan, what do you think?} 
\jy{We can revisit the results. It is not surprising that venue effect is no
  longer significant after the batch effect was included. We could do the same
  analysis with the batch-random-effect model.}

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = `ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
\eds{are you including preliminary heat data as possible within cluster 
observations in this analysis?  Based on your 
discussion in Section 2.1 about how they are not comparable, it may not be fair 
to include them here?  Or perhaps do the analysis with and without prelim heat
data to see if it makes a difference?}
\of{Yes I am including prelminary heat data.  The primary reason is that there
would be few datapoints otherwise.  I ran the numbes without prelims, the number
of observations dropped from 65 to 26 and the number of clusters from 15 to 9, the
p value increased from 0.000155 to 0.007332.
The calculations are at the bottom of RankComparison.Rmd}

@@

\jy{motivate first why you want to run the analysis with and without the 2022 data.}

@@

\jy{Close captions with a period.}
\of{Resolved}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Old Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Old Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.08, and
  0.08 for various data sets.}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   Old Finals & $4.36\cdot10^{-4}$ & $3.0\cdot10^{-7}$ \\
   All Finals & $8.64\cdot10^{-4}$ & $6.0\cdot10^{-7}$ \\
   Old Pooled & $4.15\cdot10^{-3}$ & $4.67\cdot10^{-5}$ \\ 
   All Pooled & $5.15\cdot10^{-3}$ & $6.01\cdot10^{-5}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

@@

\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Compares the various effect models by dataset, all probabilities are
  for observing a reaction time less than 0.1 seconds. In general the model with
  the highest probability (least extreme reaction time) is the Year Effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Year Effect & Batch Effect & Both Effects \\ 
   \midrule
   Old Finals & $4.39\cdot10^{-4}$ & $4.24\cdot10^{-4}$ & $4.36\cdot10^{-4}$ \\
   All Finals & $8.63\cdot10^{-4}$ & $8.65\cdot10^{-4}$ & $8.64\cdot10^{-4}$ \\
   Old Pooled & $5.73\cdot10^{-3}$ & $5.21\cdot10^{-3}$ & $4.15\cdot10^{-3}$ \\ 
   All Pooled & $6.84\cdot10^{-3}$ & $6.64\cdot10^{-3}$ & $5.15\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:EffectComparison}
\end{table}

@@

The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and can be found in 
section~\ref{sec:results2019vs2022} and secondly comparing 2022 national
championships to the 2022 World Championships and can be found in 
~section\ref{sec:resultsnational}.
\jy{Remind the readers what the analysis was about: two-group comparison with
  group being ....}
\of{Better?}

\subsubsection{2019 World Track and Field Championships} \label{sec:results2019vs2022}
The p value from the two group comparison for 2019 and 2022 was 0.001464.  This 
value is for a one-sided test and show that the mean reaction time on average in
2019 was larger (slower) than the mean reaction time in 2022 for athletes who 
competed in both championships at an $\alpha$ level of 0.01. This provides good
evidence that on average, the reaction times from the 2019 and 2022 World 
Championships are significantly different.  
\eds{That is not the only explanation} \of{You are correct so I changed the 
wording}

\subsubsection{2022 National Championships} \label{sec:resultsnational}
The p value from the two group comparison between the 2022 national
championships and World Championships was 0.000155. This value is for a one sided
test and the result is significant at $\alpha = 0.01$ and shows that
the times at the World Championships were significantly faster than at the
national championships.

\section{Discussion}\label{sec:Discussion}
One conclusion from the analysis in \citet{zhang2021correlation} was that 
athletes' reaction times improved from 2011 to 2019.  However this study does 
not consider data from 1999 to 2009, and over this timespan athletes get slower 
and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
reject's Zhang et al.'s conclusion that athletes are getting faster because they 
were fast over twenty years ago. One possible explanation for why 2011 had such
high variation was because of the IAAF rule discussed earlier that increased the
penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
Athletes may have been overly cautious to false start and may have consciously
or sub-consciously reacted slower as a result.  Instead, a more appropriate 
conclusion may be that there is a lot of variability in reaction times and that
the context of the races matters.

>>>>

\usepackage{amsmath, amsfonts}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{cleveref}
\hypersetup{colorlinks = true, linkcolor = blue, citecolor=blue, urlcolor = blue}
\usepackage{natbib}
\usepackage{float}
\usepackage{setspace}
\usepackage{pdfpages}
\usepackage[pagewise]{lineno}
\usepackage{mwe}
\usepackage{comment}
\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.

\jy{Use double blank lines to separate paragraphs in tex source. Acknowledge if you copy.}

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Unfortunately, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
It is unknown how many athletes since the 2009 study could have potentially
benefited from this, but it is probable that there are others besides Allen who
were disqualified. Additionally,
\citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year) and a second random effect accounting for the
heat effect within each year. \jy{do a thorough check to make sure this is updated.}
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

@@

\jy{give specifics of your sample size}

\begin{figure}[tbp]
  \centering
  \includegraphics{Scatterplot2019vs2022}
  \caption{The reaction times from athletes who competed at least once in both 
  the 2019 and 2022 World Track and Field Championships}
  \label{fig:Scatterplot2019vs2022}
\end{figure}
\of{I included this here for now, but feel free to change the location.  I will
add references to it and explanation of it along with some descriptive statistics about
the size and mean/median,std dev, size of the data.}

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}.

With our small sample size, the asymptotic normal distribution may not be
reliable, so we use random permutation to obtain the p-value of the observed
test statistics $W^*$. This method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = `ds'} and \texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

\jy{Use the pooled data; present the analysis with 2022 first; then suggest that
  is worth looking at the results without 2022.}

\jy{Write the story up as if you are going to submit. Then I'll review it.}

@@

\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Old Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Old Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.08, and
  0.08 for various data sets.}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   Old Finals & $4.36\cdot10^{-4}$ & $3.0\cdot10^{-7}$ \\
   All Finals & $8.64\cdot10^{-4}$ & $6.0\cdot10^{-7}$ \\
   Old Pooled & $4.15\cdot10^{-3}$ & $4.67\cdot10^{-5}$ \\ 
   All Pooled & $5.15\cdot10^{-3}$ & $6.01\cdot10^{-5}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

@@

\begin{table}
  \centering
  \caption{Compares the various effect models by dataset, all probabilities are
  for observing a reaction time less than 0.1 seconds. In general the model with
  the highest probability (least extreme reaction time) is the Year Effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Year Effect & Batch Effect & Both Effects \\ 
   \midrule
   Old Finals & $4.39\cdot10^{-4}$ & $4.24\cdot10^{-4}$ & $4.36\cdot10^{-4}$ \\
   All Finals & $8.63\cdot10^{-4}$ & $8.65\cdot10^{-4}$ & $8.64\cdot10^{-4}$ \\
   Old Pooled & $5.73\cdot10^{-3}$ & $5.21\cdot10^{-3}$ & $4.15\cdot10^{-3}$ \\ 
   All Pooled & $6.84\cdot10^{-3}$ & $6.64\cdot10^{-3}$ & $5.15\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:EffectComparison}
\end{table}

@@

The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and can be found in 
Section~\ref{sec:results2019vs2022} and secondly comparing 2022 national
championships to the 2022 World Championships and can be found in
Section~\ref{sec:resultsnational}.

\jy{use two-sided}

\jy{include sample sizes; permutation number (B);}

The p value from the two group comparison for 2019 and 2022 was 0.001464.  This 
value is for a one-sided test and show that the mean reaction time on average in
2019 was larger (slower) than the mean reaction time in 2022 for athletes who 
competed in both championships at an $\alpha$ level of 0.01. This provides good
evidence that on average, the reaction times from the 2019 and 2022 World 
Championships are significantly different.  
\jy{elaborate on the inprepretation; what we have rejected;}

The p value from the two group comparison between the 2022 national
championships and World Championships was 0.000155. This value is for a one sided
test and the result is significant at $\alpha = 0.01$ and shows that
the times at the World Championships were significantly faster than at the
national championships.

\section{Discussion}\label{sec:Discussion}

\jy{condense; could use chatGPT to improve your writing.}

One conclusion from the analysis in \citet{zhang2021correlation} was that 
athletes' reaction times improved from 2011 to 2019.  However this study does 
not consider data from 1999 to 2009, and over this timespan athletes get slower 
and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
reject's Zhang et al.'s conclusion that athletes are getting faster because they 
were fast over twenty years ago. One possible explanation for why 2011 had such
high variation was because of the IAAF rule discussed earlier that increased the
penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
Athletes may have been overly cautious to false start and may have consciously
or sub-consciously reacted slower as a result.  Instead, a more appropriate 
conclusion may be that there is a lot of variability in reaction times and that
the context of the races matters.
